Recurrent event analysis in the presence of functional covariates via random subsampling

### 2019-06-06 write-up/fda-recurrent.tex
Notes added deriving the score equations

The component of the log-likelihood related to the event process
is given by
\[
  L_n (\theta) = \sum_{i=1}^{n} \left ( \sum_{j=1}^{k_i} 
  \left[ \log ( h_i (T_{i,j}) ) - H_i (T_{i,j}, T_{i,j-1}; \theta)
  \right] - H_{i} (\tau_i, T_{i,k_i}; \theta) \right) + \log \left(
  \pr ( \bfx_i ; \psi ) \right)
\]
where~$T_{i,0} = 0$, $T_{i, k_i + 1} = \tau_i$, and
\[
H_{i} (T_{i,j}, T_{i,j-1}; \theta) = \int_{T_{i,j-1}}^{T_{i,j}} h_{i} (t; \theta) dt.
\]
Solving the associated score equations~$U_n (\theta) = 0$ yields the
maximum likelihood estimator~$\hat \theta$, where
\[
U_n (\theta) = \sum_{i=1}^{n} \left ( \sum_{j=1}^{k_i} 
  \left[ \frac{h^{(1)}_i (T_{i,j})}{h_i (T_{i,j})} - H^{(1)}_i
    (T_{i,j}, T_{i,j-1}; \theta) \right]  - H^{(1)}_{i} (\tau_i,
  T_{i,k_i}; \theta) \right) 
\]
and~$h_i^{(1)} (T_{i,j}; \theta)$ and $H_i^{(1)} (T_{i,j}, T_{i,j-1}; \theta)$
are the derivatives of~$h_i^{(1)} (T_{i,j}; \theta)$ and~$H_i^{(1)}
(T_{i,j}, T_{i,j-1}; \theta)$ with repsect to $\theta$.

In classical joint models~\citep{Henderson2000, Tsiatis2004},
time-varying covariates~$x_i (t)$ are observed only intermittently at
appointment times, which may or may not be
administrative~\citep{DempseyPMCC}.  In this setting, it is not feasible
to compute the cumulative hazard functions~$H_{i} (T_{i,j}, T_{i,j-1};
\theta)$.  In our current setting, we assume~$x_{i} (t)$ is observed
continuously.  While this relieves the issue of intermittent
observations, a joint model requires a flexible class of longitudinal
models for the sensor stream data that adequately capture their
dynamics.  Moreover, the risk model now depends on an integrated past
history of the time-varying covariate leads to severe increase in
computational complexity.

@@

An estimator is design-unbiased if its expectation is equal to that
parameter under the probability distribution induced by the sampling
design~\citep{Cassel1977}. Let~$D_i \subset [0,t_i]$ denote the set of
subsampled points.  Note, by design, this set is distinct from the set
of event times, i.e.,~$\bfT_i \cap D_i = \emptyset$.  Under
subsampling via $\pi_i (t)$, then we may compute the Horvitz-Thompson
estimator 
\[
\hat H_{i} (T_{i,j}, T_{i,j-1}; \theta) = \sum_{t \in D \cap [T_{i,j-1},
  T_{i,j}]} \frac{ \lambda^{(1)} ( t \given \{ x(t - s) \}_{0 < s <
    \Delta}; \theta ) }{ \pi (s) }
\]
Let~$N = \{ T_{i,j} \}_{i =1, j=1}^{n,k_i}$ denote the complete set of
event times.  Then an alternative design-unbiased estimator of the
cumulative hazards is given by
\begin{equation}
\label{eq:WPest}
\hat H_{i} (T_{i,j}, T_{i,j-1}; \theta) = \sum_{t \in (N \cup D) \cap [T_{i,j-1},
  T_{i,j}]} \frac{ \lambda^{(1)} ( t \given \{ x(t - s) \}_{0 < s <
    \Delta}; \theta ) }{ \pi (s) + \lambda ( t \given \{ x(t - s) \}_{0 < s <
    \Delta}; \theta) }
\end{equation}
Equation~\eqref{eq:WPest} is the estimator suggested
by~\cite{Waagepetersen2008}.  This estimator depends on the
superposition of the event and subsampling processes.  This estimator
has been shown to be more efficient than the Horvitz-Thompson
estimator; therefore, we restrict our attention to~\eqref{eq:WPest}
for the remainder of this paper. Let
\[
w (t; \theta) = \frac{\pi (t)}{\pi (t) + \lambda (t \given \{
  x(t - s) \}_{0 < s < \Delta}; \theta)}
\]
Then the resulting estimating equations are given by
\begin{equation}
\label{eq:approxscore}
\hat{U}_n = \sum_{t \in N} w(t; \theta) \frac{\lambda^{(1)} (t \given \{
  x(t - s) \}_{0 < s < \Delta}; \theta)}{ \lambda ( t \given \{
  x(t - s) \}_{0 < s < \Delta}; \theta)}  - 
\sum_{t \in D} w(t; \theta) \frac{\lambda^{(1)} (t \given \{
  x(t - s) \}_{0 < s < \Delta}; \theta)}{ \pi (t) }
\end{equation}
Equation~\eqref{eq:approxscore} are approximate score functions built
via plug-in of the design-unbiased estimator of the cumulative hazard.

@@

\section{Functional principal components within event-history
  analysis}

@@

A simpler alternative is the Horvitz-Thompson estimator where the
denominator only depends on

\[
L(\theta) = \sum_{s \in N} \log \lambda \left( \{ x(s-t) \}_{0 < t
    \leq \Delta} \right) 
\]

@@

@@

\subsection{}

>>>>

For the sake of notational simplicity, we leave the dependency of the
conditional hazard function on~$H_{i,t}^{NX}$ implicit, and write~$h_i
(t; \theta)$. The component of the log-likelihood related to the event
process is then given by
\[
  L_n (\theta) = \sum_{i=1}^{n} \left ( \sum_{j=1}^{k_i} 
  \left[ \log \left( h_i \left( T_{i,j}; \theta
      \right) \right) - H_i \left( T_{i,j}, T_{i,j-1}; \theta \right) 
  \right] - H_{i} \left( \tau_i, T_{i,k_i}; \theta \right) \right) 
\]
where~$T_{i,0} = 0$, $T_{i, k_i + 1} = \tau_i$, and
\[
H_{i} (T_{i,j}, T_{i,j-1}; \theta) = \int_{T_{i,j-1}}^{T_{i,j}} h_{i} (t; \theta) dt.
\]
Solving the associated score equations~$U_n (\theta) = 0$ yields the
maximum likelihood estimator~$\hat \theta$, where
\[
U_n (\theta) = \sum_{i=1}^{n} \left ( \sum_{j=1}^{k_i} 
  \left[ \frac{h^{(1)}_i (T_{i,j}; \theta)}{h_i (T_{i,j}; \theta)} - H^{(1)}_i
    (T_{i,j}, T_{i,j-1}; \theta) \right]  - H^{(1)}_{i} (\tau_i,
  T_{i,k_i}; \theta) \right) 
\]
and~$h_i^{(1)} (T_{i,j}; \theta)$ and $H_i^{(1)} (T_{i,j}, T_{i,j-1}; \theta)$
are the derivatives of~$h_i^{(1)} (T_{i,j}; \theta)$ and~$H_i^{(1)}
(T_{i,j}, T_{i,j-1}; \theta)$ with respect to $\theta$.

In classical joint models~\citep{Henderson2000, Tsiatis2004},
time-varying covariates~$x_i (t)$ are observed only intermittently at
appointment times. In our current setting, maximizing the likelihood
is computationally prohibitive since for any~$\theta$ we must compute
the cumulative hazard functions~$H_{i} (T_{i,j}, T_{i,j-1}; \theta)$
which require integration of~$h_i(t;\theta)$ given
by~\eqref{eq:hazardlinear} which itself depends on the
integral~$\int_{t-\Delta}^t x_i (s) \beta(s) ds$.  
That is, the risk model now depends on an integrated past history of
the time-varying covariate which leads to severe increase in
computational complexity.

@@

An estimator is design-unbiased if its expectation is equal to that
parameter under the probability distribution induced by the sampling
design~\citep{Cassel1977}. Let~$D_i \subset [0,t_i]$ denote the set of
subsampled points.  Note, by design, this set is distinct from the set
of event times, i.e.,~$\bfT_i \cap D_i = \emptyset$.  Under
subsampling via $\pi_i (t)$, then we may compute the Horvitz-Thompson
estimator 
\[
\hat H_{i} (T_{i,j}, T_{i,j-1}; \theta) = \sum_{t \in D_i \cap [T_{i,j-1},
  T_{i,j}]} \frac{ \lambda^{(1)} ( t \given \{ x(t - s) \}_{0 < s <
    \Delta}; \theta ) }{ \pi (s) }
\]
Let~$N = \{ T_{i,j} \}_{i =1, j=1}^{n,k_i}$ denote the complete set of
event times.  Then an alternative design-unbiased estimator of the
cumulative hazards is given by
\begin{equation}
\label{eq:WPest}
\hat H_{i} (T_{i,j}, T_{i,j-1}; \theta) = \sum_{t \in (N \cup D) \cap [T_{i,j-1},
  T_{i,j}]} \frac{ \lambda^{(1)} ( t \given \{ x(t - s) \}_{0 < s <
    \Delta}; \theta ) }{ \pi (s) + \lambda ( t \given \{ x(t - s) \}_{0 < s <
    \Delta}; \theta) }
\end{equation}
Equation~\eqref{eq:WPest} is the estimator suggested
by~\cite{Waagepetersen2008}.  This estimator depends on the
superposition of the event and subsampling processes.  This estimator
has been shown to be more efficient than the Horvitz-Thompson
estimator; therefore, we restrict our attention to~\eqref{eq:WPest}
for the remainder of this paper. Let
\[
w (t; \theta) = \frac{\pi (t)}{\pi (t) + \lambda (t \given \{
  x(t - s) \}_{0 < s < \Delta}; \theta)}
\]
Then the resulting estimating equations are given by
\begin{equation}
\label{eq:approxscore}
\hat{U}_n = \sum_{t \in N} w(t; \theta) \frac{\lambda^{(1)} (t \given \{
  x(t - s) \}_{0 < s < \Delta}; \theta)}{ \lambda ( t \given \{
  x(t - s) \}_{0 < s < \Delta}; \theta)}  - 
\sum_{t \in D} w(t; \theta) \frac{\lambda^{(1)} (t \given \{
  x(t - s) \}_{0 < s < \Delta}; \theta)}{ \pi (t) }
\end{equation}
Equation~\eqref{eq:approxscore} are approximate score functions built
via plug-in of the design-unbiased estimator of the cumulative hazard.

@@

\section{Longitudinal functional principal components within
  event-history analysis}

Probabilistic subsampling converts the single sensor stream~$\bar x_i$
into a sequence of functions observed repeatedly at sampled
times~$D_i$ and event times~$N_i$. Such a data structure is commonly
referred to as \emph{longitudinal functional data}~\ref{Xiao2013,
  GoldSmith2015}.
Given the large increase in longitudinal functional data in recent
years, corresponding analysis has received much recent
attention~\citep{Morris2003, MorrisCarroll2006,
  Baladandayuthapani2008, Di2009, Greven2010, Staicu2010,
  ChenMuller2012, LiGuan2014}.
In this paper, we combine work by~\cite{Park2018}
and~\cite{Goldsmith2011} to construct a computationally efficient
penalized functional method for solving the estimation equations~$\hat
U_n (\theta)$.

In our setting, probabilistic subsampling leads to observation
times~$t \in N \cup D$ at which

The marginal covariance~\cite{Park2018} allows a single 
\begin{itemize}
\item {\bf Step 1}: Estimate~$\hat \mu(s,t)$ via sandwich smoother for
  all~$s \in [0,\Delta]$ and~$t \in \cup_{i=1}^n \left( D_i \cup N_i
  \right)$. Define~$\tilde X (s,t) = X(s,t) - \hat \mu(s,t)$ to be a
  mean-zero process for each~$t$.
\item {\bf Step 2}: Compute the pooled sample covariance
\[
\hat \Xi (s , s^\prime) = 
\left( \sum_{i=1}^n \left | N_i \cup D_i \right |\right)^{-1}
\left( \sum_{i=1}^n \sum_{t \in N_i \cup D_i} \tilde X (s,t) \tilde X
  (s^\prime, t) \right).
\]
Compute at a fine set of grid points. This is an estimate of 
\[
\Xi (s, s^\prime) = \int_{\mathcal{T}} c( (s,T), (s^\prime, T) ) g(T) dT
\]
where~$g(T)$ is a superposition of the event process and the
subsampling process.
\item {\bf Step 3}: Take spectral decomposition~$\{ \hat \phi_k (s),
  \hat \lambda_k \}_{k \geq 1}$. Choose~$K_x \geq K_b$ large enough.
\item {\bf Step 4}: Compute
\[
\hat c_{i,k} (t) = \int_{t-\Delta}^t \tilde X_i (s,t) \hat \phi_k (s) ds
\]
Then
\[
X(s,t) \approx \hat \mu(s,t) + \sum_{k=1}^{K_x} \hat c_{i,k} (t)
\phi_k (s).
\]
\item {\bf Step 5}: Plugging both the spline and KL decomposition we
  have
\begin{align*}
\int_{t-\Delta}^t X(s,t) \beta(s) ds 
  &= \int_{t-\Delta}^t \left[ \hat \mu(s,t) + \sum_{k=1}^{K_x}
    c_{i,t,k} \phi_k (s) \right] \times \left[ \sum_{k^\prime=1}^{K_b}
    \beta_k \psi_k (s) \right] \\
  &= [ M_{i,t}^\prime + C_{i,t}^\prime J_{\phi, \psi} ] \beta
\end{align*}
where~$M_{i,t}$ is a vector with~$i$th entry equal
to~$\int_{t-\Delta}^t \hat \mu(s,t) \psi_k (s) ds$, $C_{i,t} =
(c_{i,t,1}, \ldots, c_{i,t,K_x}) \in \mathbb{R}^{K_x \times 1}$,
$\beta = (\beta_1, \ldots, \beta_{K_b}) \in \mathbb{R}^{K_b \times 1}$
and~$J_{\phi, \psi} \in \mathbb{R}^{K_x \times K_b}$ where the~$(i,j)$
entry is given by
\[
\int_{t-\Delta}^t \phi_{i} (s) \psi_{j} (s) ds.
\]
\end{itemize}

We re-write the hazard function as:
\[
h_i (t; \theta) = \exp \left( Z_{i,t}^\prime, \alpha + [M_{i,t} + C_{i,t}^\prime
  J_{\phi, \psi} ] \beta \right) = \exp \left( W_{i,t}^\prime \theta \right),
\]
where~$W_{i,t}^\prime = [ Z_{i,t}^\prime (M_{i,t}^\prime +
C_{i,t}^\prime J_{\phi, \psi}) ]$ and~$\theta = (\alpha,
\beta)$. Then, the score function is:
\[
\hat U_n (\theta) = \sum_{i=1}^n \left \{  \sum_{t \in N_i}
  \frac{h_i^{(1)} (t; \theta)}{h_i (t; \theta)} - \hat H_i^{(1)}
  (\tau_i, 0; \theta) 
\right \}
\]
where
\[
\hat H_i^{(1)} (\tau_i; \theta) = \sum_{u \in N_i \cup D_i} \frac{
  h^{(1)}_i (u; \theta)}{\pi_i (u) + h_i (u; \theta)}.
\]
Note that
\[
\frac{d}{d\theta} \log h_i (t; \theta) = \frac{h^{(1)}_i (t;
  \theta)}{h_i (t; \theta)}.
\]
Therefore, these terms have very simple forms!!! 
\[
\frac{d}{d\theta} \log h_i (t; \theta) = W_{i,t}
\]
On the other hand,
\[
\frac{d}{d \theta} \log \left( \pi_i (t) + h_i (t;\theta) \right) = 
\frac{\exp \left( W_{i,t}^\prime \theta \right)}{\pi_i (t) + \exp
  \left( W_{i,t}^\prime \theta \right)} W_{i,t}
\]
Therefore,
\begin{align*}
\hat U_n (\theta) 
  &= \sum_{i=1}^n \left \{ \sum_{t \in N_i} W_{i,t} -
    \sum_{u \in N_i \cup D_i} \frac{\exp \left( W_{i,u}^\prime \theta
    \right)}{\pi_i (t) + \exp \left( W_{i,u}^\prime \theta \right)}
    W_{i,u} \right \} \\
  &= \sum_{i=1}^n \left \{ \sum_{t \in N_i} \frac{\pi_i (t)}{\pi_i (t)
    + \exp \left( W_{i,u}^\prime \theta \right)} W_{i,t} -
    \sum_{u \in D_i} \frac{\exp \left( W_{i,u}^\prime \theta
    \right)}{\pi_i (t) + \exp \left( W_{i,u}^\prime \theta \right)}
    W_{i,u} \right \} \\
  &= \sum_{i=1}^n \left \{ \sum_{t \in N_i} w_{i} (t; \theta) W_{i,t} -
    \sum_{u \in D_i} (1 - w_i (u; \theta)) W_{i,u} \right \} \\
\end{align*}

\section{Prediction}
We wish to estimate the risk at non-sampled points.

For prediction, we use sparse FPCA (Yao 2005) 
\begin{itemize}
\item Take $\hat c_{k} (s,t)$ for~$t \in N \cup D$ and~$s \in
  [0,\Delta]$
\item 
\end{itemize}

@@

\[
- b^\prime D b
\]

@@

\begin{thm}[No reconstruction error]

\end{thm}

\begin{thm}[Reconstruction error but~$\hat \Delta > \Delta$]
\label{eq:biaseddown}
\end{thm}

Theorem~\ref{eq:}

\begin{thm}[When~$\hat \Delta < \Delta$]

\end{thm}

@@

\subsection{Unbiased estimation}

\subsection{Unbiased estimation}
